---
title: "earthquakesNOAA"
author: "Williams Ojo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

National Oceanographic and Atmospheric Administration (NOAA) collects significant earthquakes around the world and make a dataset that contains information about 5,933 earthquakes over an approximately 4,000 year time span. This dataset has a substantial amount of information embedded in it that may not be immediately accessible to people without knowledge of the intimate details of the dataset.

The overall goal of the R package is to work with this dataset and provide the tools for processing and visualizing the data so that others may extract some use out of the information embedded within.

```{r, eval=FALSE}
library(devtools)
devtools::install_github('willojs/earthquakesNOAA', build_vignettes = TRUE)
library(earthquakesNOAA)
```

## Examples

Download the data from the NOAA website, saved it to your working directory and transform it to a data frame using the `read_delim` function:
  
Clean data
```{r eval = FALSE}
filename <- system.file("extdata","earthquake.txt", package="earthquakesNOAA")
library(readr)
data <- readr::read_delim(filename, delim = "\t")
```

Before using the visualization tools, the data must be cleaned with the functions `eq_clean_data` and `eq_location_clean`.

```{r eval = FALSE}
clean_data <- eq_clean_data(data)
clean_data <- eq_location_clean(clean_data)
```

To map the earthquakes epicenters in Mexico since 2000 and providing their dates in annotation use the `eq_map` function:

```{r eval = FALSE}
clean_data %>%
  dplyr::filter(COUNTRY == "MEXICO" & lubridate::year(DATE) >= 2000) %>%
  eq_map(annot_col = "DATE")
```

To show annotations, use the `eq_create_label` function before the `eq_map` function:

```{r eval = FALSE}
clean_data %>%
  dplyr::filter(COUNTRY == "MEXICO" & lubridate::year(DATE) >= 2000) %>%
  dplyr::mutate(popup_text = eq_create_label(.)) %>%
  eq_map(annot_col = "popup_text")
```

To visualize the times, the magnitudes and the number of deaths associated to earthquakes within certain countries, use the `geom_timeline` geom with the `ggplot` function:

```{r eval = FALSE}
data %>%
  dplyr::filter(COUNTRY == c("MEXICO","USA") & lubridate::year(DATE) >= 2010) %>%
  ggplot(aes(x=DATE,y=COUNTRY,color=TOTAL_DEATHS,size=EQ_PRIMARY)) +
  geom_timeline(alpha=.5) +
  theme(legend.position="bottom", legend.box="horizontal", plot.title=element_text(hjust=0.5)) +
  ggtitle("Earthquakes Visualization Tool") +
  labs(size = "Richter scale value", color = "# deaths")
```

Use the `geom_timelinelabel` geom for adding annotations to the earthquake data and an option to subset the n_max largest earthquakes by magnitude:

```{r eval = FALSE}
data %>%
  dplyr::filter(COUNTRY == c("MEXICO","USA") & lubridate::year(DATE) >= 2010) %>%
  ggplot(aes(x=DATE,y=COUNTRY,color=TOTAL_DEATHS,size=EQ_PRIMARY)) +
  geom_timeline(alpha=.5) +
  geom_timelinelabel(aes(label=LOCATION_NAME),n_max=3) +
  theme(legend.position="bottom", legend.box="horizontal", plot.title=element_text(hjust=0.5)) +
  ggtitle("Earthquakes Visualization Tool") +
  labs(size = "Richter scale value", color = "# deaths")
