---
title: "earthquakesNOAA"
author: "Williams Ojo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The package 'earthquakeNOAA' has been built to clean and display the earthquake data from NOAA (The National Centers for Environmental
Information). The dataset can be found here: https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1

The Package is divided into 3 sections:

* Cleaning the Dataset
* Displaying the Earthquakes through the geom_timeline and geom_timeline_label functions.
* Displaying the Earthquakes by their geographical location, through leaflet.

These sections give you a comprehensive way to analyze and visualize the Earthquake data. 

## Cleaning the Dataset

The cleaning part of the package is composed using 2 functions:

### eq_location clean

'eq_location_clean' takes an argument 'data', and returns that same data with its column LOCATION_NAME cleaned.
The cleaning is done by splitting the character string by the colon ':' symbol. Then, the latest of the strings is selected (
the rigth-most string). Finally, the strings are converted to title case, with the first letter in caps.

```{r, echo = FALSE}
if (require(earthquakesNOAA)){
  library(earthquakesNOAA)
}
```

```{r, warning=FALSE}
## Data Setup
test_df = data.frame(order = 1:10,
                            LOCATION_NAME = c("Peru: Lima", "Mexico: Tijuana",
                                              "CUBA: HABANA", "SRI LANKA: IslanD",
                                              "EU: SPain: unknown", "US: Maine",
                                              "US: California: Los Angeles", " Ecuador: Bogota",
                                              "Ivory COAST: ISLAND 3", "China: HONG KONG"),
                            stringsAsFactors = F)
## Function Call
test_df = eq_location_clean(test_df)
test_df
  
```

### eq_clean_data

'eq_clean_data' takes an argument 'data', same as 'eq_location_clean', and returns a cleaned dataset with the 'eq_location_clean' applied to it. Furthermore, the LATITUDE, LONGITUDE, DEATHS, TOTAL_DEATHS columns are converted to numeric type. The data points with YEAR columns below 0 (BC years) are removed. The NA values in DAY, and MONTH columns are  converted to 1. Finally, a new DATE column is created by combining the DAY, MONTH, and YEAR columns and converting them to Date type.


```{r}
## Data Setup
test_df = data.frame(LOCATION_NAME = c("Peru: Lima", "Mexico: Tijuana",
                                              "CUBA: HABANA", "SRI LANKA: IslanD",
                                              "EU: SPain: unknown", "US: Maine",
                                              "US: California: Los Angeles"),
                            YEAR = c(2150,2030,2000,1999,-1610,1566,1450),
                            MONTH = c(NA,3,NA,1,9,6,5),
                            DAY = c(NA,NA,1,4,NA,2,NA),
                            LATITUDE = c(-12.5,20.2,34.1,8.567,36.5,46.975,34.262),
                            LONGITUDE = c(-77,-103.5,-75.5,81.233,-4.4,-66.659,-118.002),
                            DEATHS = c(NA,1,2,8,33,NA,NA),
                            TOTAL_DEATHS = c(NA,1,NA,9,NA,25,NA),
                            stringsAsFactors = F)
## Function Call
test_df = eq_clean_data(test_df) # Calling the function
test_df
```


## GeomTimeline & GeomTimelineLabel

The newly created geoms, Geom Timeline & Geom Timeline Label, display the earthquake data through points of varying color, size, and location (based on coordinates).

### GeomTimeline

'GeomTimeline' is a new Geom made of pointGrobs and lineGrobs, which display the input 'data' as points. These points vary based on the colour and size arguments. If a y-axis is given, the points will be displayed in different y-coordinates. Furthermore, a straight line is drawn from the minimum x-value to the maximum x-value. If there are y-coordinates, the line is drawn for each y. This Geom works once it is called by its  function 'geom_timeline'.

### geom_timeline

'geom_timeline' is the layer function for the GeomTimeline geom. To call the Geom, the 'data' argument must be supplied.
Furthermore, the colour and size arguments will indicate the color and size of the data points. If, the y argument is supplied,
the data points points will be grouped into n(y) groups, each displaying its own line of points.


```{r, fig.width=7.2, fig.height=3}
## Data Setup
test_df = data.frame(LOCATION_NAME = c("Peru: Lima", "Mexico: Tijuana",
                                              "CUBA: HABANA", "SRI LANKA: IslanD",
                                              "EU: SPain: unknown", "US: Maine",
                                              "US: California: Los Angeles"),
                            YEAR = c(2150,2030,2000,1999,-1610,1566,1450),
                            MONTH = c(NA,3,NA,1,9,6,5),
                            DAY = c(NA,NA,1,4,NA,2,NA),
                            LATITUDE = c(-12.5,20.2,34.1,8.567,36.5,46.975,34.262),
                            LONGITUDE = c(-77,-103.5,-75.5,81.233,-4.4,-66.659,-118.002),
                            DEATHS = c(NA,1,2,8,33,NA,NA),
                            TOTAL_DEATHS = c(NA,1,NA,9,NA,25,NA),
                            EQ_PRIMARY = c(4.4,6.5, 8.1, 2.7, 5.5, 6.3, 7.4),
                            stringsAsFactors = F)
test_df = eq_clean_data(test_df)
## Function Call
ggplot2::ggplot(test_df, ggplot2::aes(DATE,
                                      colour = TOTAL_DEATHS,
                                      size = as.numeric(EQ_PRIMARY) )) +
  geom_timeline() +
  ggplot2::guides(size = ggplot2::guide_legend(title = "Ritcher scale value")) +
  ggplot2::scale_colour_continuous(name = "# of DEATHS") +
  ggplot2::theme_classic()
```


### GeomTimelineLabel

'GeomTimelineLabel' is a new geom made of lineGrobs and textGrobs, which label the points from the `data` argument. The location of these labels are based on the provided 'x' argument. The text used to label should be given as the `label` argument. The textGrob is positioned above the x-coordinate of the points, right above a vertical line drawn at the point. If y-coordinates are given, the vertical lines drawn will be smaller in size, depending on how many unique y-coordinates there are.

### geom_timeline_label

'geom_timeline_label' is the layer function to call the GeomTimelineLabel geom. This label works perfectly with the previously defined 'geom_timeline' function, and inherits its parameters as with any other ggplot. The extra argument needed is the 'label' argument. Another optional argument used is 'nmax', which describes how many top labels to use in the annotation. By 'top' label, we ferer to the data points with the higher 'nmax' values based on their `size` attribute. Note: If you need to provide a size, nmax, or other argument as a single value, rather than a column of the data, provide it outside the aes function.


```{r, fig.width=7.2, fig.height=3}
## Function Call
ggplot2::ggplot(test_df, ggplot2::aes(DATE,
                                      colour = TOTAL_DEATHS,
                                      size = as.numeric(EQ_PRIMARY) )) +
  geom_timeline() +
  geom_timeline_label(ggplot2::aes(label = test_df$LOCATION_NAME)) + # Labeling
  ggplot2::guides(size = ggplot2::guide_legend(title = "Ritcher scale value")) +
  ggplot2::scale_colour_continuous(name = "# of DEATHS") +
  ggplot2::theme_classic()
```


## Leaflet Functions

The following 2 functions were built to display the earthquake data by their geographical location through a leaflet map.

### eq_map

`eq_map` is a function that displays the data points in the input `data` argument, based on its LATITUDE and LONGITUDE columns. The data points will be displayed as circles in the map. The second argument, `annot_col` decides which column in the dataset will be used as the pop-up annotation of the leaflet map.


**Note: Leaflet Maps may not display in some browsers. To visualize the results of the leaflet map functions (in vignettes), download the package and knit the vignette (or preview the html file) in RStudio.**
```{r, fig.width=7.2, fig.height=3}
## Data Setup
## This is test data (not actual NOAA Locations)
test_df = data.frame(LOCATION_NAME = c("Mexico: Jalisco",
                                              "MEXICO:  OAXACA",
                                              "US: California: Los Angeles"),
                            YEAR = c(2000,1610,1450),
                            MONTH = c(3,NA,5),
                            DAY = c(1,4,NA),
                            LATITUDE = c(20.2,17.0,34.262),
                            LONGITUDE = c(-103.5,-96.3,-118.002),
                            DEATHS = c(1,2,NA),
                            TOTAL_DEATHS = c(1,NA,25),
                            EQ_PRIMARY = c(6.5, 2.7, 7.4),
                            stringsAsFactors = F)
test_df = eq_clean_data(test_df)
## Function Call
m = eq_clean_data(test_df)
eq_map(m, annot_col = "DATE")
```

### eq_create_label

`eq_create_label` creates a new column called 'popup_text' from the LOCATION_NAME, EQ_PRIMARY, and TOTAL_DEATHS columns of the input argument `data`. The resulting column is a character vector formatted in HTML-format, where the column LOCATION_NAME is titled 'Location', the column 'EQ_PRIMARY' is titled 'Magnitude', and the column 'TOTAL_DEATHS' is titled 'Total deaths'. If any of the dataset's rows is missing either of the column values, then that particular annotation is skipped in the html string.


```{r, fig.width=7.2, fig.height=3}
## Function Call 
m$popup_text = eq_create_label(m) # Creating the Labels
## The locations and coordinates are random (not actual lat and long)
eq_map(m, annot_col = "popup_text")
```
